---
title: "Kaggle Project: House Prices"
author: "Shreyas"
date: "12/05/2021"
knit: (function(inputFile, encoding) { 
      out_dir <- 'docs';
      rmarkdown::render(inputFile,
                        encoding=encoding, 
                        output_file=file.path(dirname(inputFile), out_dir, 'Kaggle_House_Prices.html')) })
output: 
  html_document:
      toc: true
---

# Project Approach

> The goal of this project is to predict house prices. The approach which we used begins with Exploratory Data Analysis (EDA). We then used Data visualization and plots with ggplot to study the data. We cleaned the data and preprocessed it to create a model. We used SalePrice as our main predictor and then tried to find correlated variables to see what variables would be useful in predicting house prices. We narrowed it down to the best variables and then we used a linear regression model to predict the home prices with the variables that add value to home prices. To verify our results, we cross-validated our predicted values, which allowed us to see if the results were consistent among the different models employed. 


```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE) 
```

## Libraries

### Loading the required libraries

>This code will install the missing libraries and then load them.

```{r libraries}

packages <- c("ggplot2","dplyr","plyr","tidyverse","data.table","knitr","kableExtra","caret","corrplot","randomForest","Hmisc","moments","psych","glmnet","scales","ggrepel","cowplot")

for(pkg in packages){
  if (pkg %in% rownames(installed.packages()) == FALSE)
  {install.packages(pkg)}
  if (pkg %in% rownames(.packages()) == FALSE)
  {library(pkg, character.only = TRUE)}
}
```

## Importing data

>We are using `read.csv` to read data into train and test variables, setting character variables to be loaded as characters instead of factors.

```{r imprting_data}

train <- read.csv("ames_train.csv", stringsAsFactors = FALSE)
test <- read.csv("ames_test.csv", stringsAsFactors = FALSE)

dim(train)
dim(test)
    
glimpse(train)

```

## Combining datasets

>  Adding SalePrice column in the test dataset to make the variables equal in both train and test datasets. 

```{r combining_data}

test$SalePrice <- NA

complete_data_set <- rbind(train, test) #Binding train & test data

complete_data_set <- cbind(complete_data_set, Set = c(rep("Train", times = dim(train)[1]),
                                rep("Test", times = dim(test)[1]))) 


dim(complete_data_set)
```

>Checking the summary of all the variables like mean, median and sd

```{r}
describe(complete_data_set, fast=TRUE)
```

## Data Exploration and cleaning

```{r}
#To check the class (character, numeric, integer) of the variables in the data set

table(sapply(complete_data_set, class)) 

```
> The dataset consists of 43 characters and 37 integer variables, and 35 columns with missing values (NA values). 


## Response variable - SalePrice

```{r SalePrice}

# Building a dataset with response variable SalePrice and its log transformation

data <- complete_data_set %>%
          select("SalePrice") %>%
          na.omit %>%
          mutate(LogSalePrice = log(SalePrice+1))

summary(data$SalePrice)

cat("Skewness: ",skewness(data$SalePrice),"\n")
cat("Kurtosis: ",kurtosis(data$SalePrice))

```
> A positive skew indicates that the tail is on the right side of the distribution, which extends towards more positive values and a kurtosis greater than 3, it is said to be `leptokurtic`, which means it tends to produce more outliers than the normal distribution. To correct this we have used the log transformation of the response variable.

## Histogram plots of SalePrice

```{r histogram_plot}

#Histogram plots of SalePrice and its log transformation

hist1 <- ggplot(data, aes(x = SalePrice, fill = ..count..), add=TRUE) +
  geom_histogram(binwidth = 5000) +
  ggtitle("Plot 1: Histogram plot of SalePrice") +
  ylab("Frequency (Number of houses)") +
  xlab("SalePrice") + 
  theme(plot.title = element_text(hjust = 0.5))

hist2 <- ggplot(data, aes(x = LogSalePrice, fill = ..count..)) +
  geom_histogram(binwidth = 0.05) +
  ggtitle("Plot 2: Histogram plot of log(SalePrice)") +
  ylab("Frequency (Number of houses)") +
  xlab("log(SalePrice)") + 
  theme(plot.title = element_text(hjust = 0.5))

cowplot::plot_grid(hist1, hist2)

```

> Histogram plot to the left is positively skewed. Plot to the right represents the log transformation of the same response variable which follows normal distribution.

## Predictor Variables

### Checking the count of NA values

```{r na_values }

variables_with_na <- which(colSums(is.na(complete_data_set)) > 0)

sort(colSums(sapply(complete_data_set[variables_with_na], is.na)), decreasing = TRUE)

```
> We studied the NA values of all the variables above and found some discrepancies and some missing values that we corrected.

## Interpreting the NA values

```{r}

#Special Case: Basement specific variables, there is a mismatch in NA values in the below variables.

complete_data_set[ !is.na(complete_data_set$BsmtFinType1) & (   
                    is.na(complete_data_set$BsmtCond) | 
                      is.na(complete_data_set$BsmtQual) | 
                      is.na(complete_data_set$BsmtExposure) | 
                      is.na(complete_data_set$BsmtFinType2)), 
                   c('BsmtQual', 'BsmtCond', 'BsmtExposure', 'BsmtFinType1', 'BsmtFinType2')
                  ]

#Special Case: Garage specific variables, there is a mismatch in NA values, will check the entries

complete_data_set[ !is.na(complete_data_set$GarageType) & is.na(complete_data_set$GarageFinish), 
                   c('GarageCars', 'GarageArea', 'GarageType', 'GarageCond', 'GarageQual', 'GarageFinish')
                 ]


```

## Function to replace NA values


```{r replace_na }

clean_na_func <- function(data_set_name) {

  #Basement special cases, missing values
  data_set_name$BsmtFinType2[333] <- names(which.max(table(data_set_name$BsmtFinType2)))
  data_set_name$BsmtExposure[c(949, 1488, 2349)] <- names(which.max(table(data_set_name$BsmtExposure)))
  data_set_name$BsmtCond[c(2041, 2186, 2525)] <- names(which.max(table(data_set_name$BsmtCond)))
  data_set_name$BsmtQual[c(2218, 2219)] <- names(which.max(table(data_set_name$BsmtQual)))
  
  #Garage special case, missing value
  data_set_name$GarageFinish[2127] <- names(which.max(table(data_set_name$GarageFinish)))
  data_set_name$GarageCond[2127] <- names(which.max(table(data_set_name$GarageCond)))
  data_set_name$GarageQual[2127] <- names(which.max(table(data_set_name$GarageQual)))

  data_set_name$GarageCars[2577] <- 0
  data_set_name$GarageArea[2577] <- 0
  data_set_name$GarageType[2577] <- "None"
  
  #correcting the error
  data_set_name$GarageYrBlt[2593] <- 2007 #GarageYrBlt=2207 looks like a typo, YearBuilt=2006, YearRemodAdd=2007.
  

  # For below variables where an absence of a feature is NA value, replacing it with 'None'

  ColVector <- c("Alley", "BsmtQual", "BsmtExposure", "BsmtFinType1", "BsmtFinType2", "FireplaceQu", "PoolQC", "Fence",
                 "MiscFeature", "GarageType", "GarageFinish", "GarageQual", "GarageCond", "BsmtCond")
  
  data_set_name[,ColVector] <- apply(data_set_name[,ColVector], 2, 
                    function(x) {
                            replace(x, is.na(x), "None")
                            }
                    )

  
# Missing value is replaced by 0 in below Numeric Variables
  
  ColVector <- c( "MasVnrArea", "BsmtFinSF2", "BsmtUnfSF", "TotalBsmtSF", "BsmtFullBath", "BsmtHalfBath")
  
  data_set_name[,ColVector] <- apply(data_set_name[,ColVector], 2, 
                    function(x) {
                            replace(x, is.na(x), 0)
                            }
                    )

# Replacing with the maximum value (mode values) in below variables

  ColVector <- c("MSZoning", "Utilities", "Exterior1st", "Exterior2nd", "MasVnrType", "Electrical", "KitchenQual", 
                 "Functional", "SaleType")
  
  data_set_name[,ColVector] <- apply(data_set_name[,ColVector], 2, 
                    function(x) {
                            replace(x, is.na(x), names(which.max(table(x))))
                            }
                    )

# Replacing the value with the median

  ColVector <- c("LotFrontage","GarageCars", "GarageArea", "BsmtFinSF1")

  data_set_name[,ColVector] <- apply(data_set_name[,ColVector], 2, 
                    function(x) {
                            replace(x, is.na(x), median(x, na.rm = T))
                            }
                    )

# Assuming that the Garage was built the same year when the house was built to replace the NA values

  data_set_name$GarageYrBlt[is.na(complete_data_set$GarageYrBlt)] <- data_set_name$YearBuilt[is.na(complete_data_set$GarageYrBlt)]
  
return(data_set_name) 
}

complete_data_set <- clean_na_func(complete_data_set)
```

## MSZoning Variable

### Boxplot of SalePrice by MSZoning

```{r}

ggplot(data=complete_data_set[!is.na(complete_data_set$SalePrice),], aes(x=factor(MSZoning), y=SalePrice)) +
 geom_boxplot(outlier.colour = "red", outlier.shape = 1)  +
        scale_y_continuous(breaks= seq(0, 800000, by=100000), labels = comma)  + 
ggtitle("Plot 3: Box plot of SalePrice by MSZoning")

```

### Relationship with GrLivArea

```{r}
complete_data_set %>%
  select(MSZoning, GrLivArea) %>%
  group_by(MSZoning) %>%
  summarise_at(vars(GrLivArea), list(count = length,avg_houseSize = mean))

```
> Average size of a house in commercial is much smaller than that in village zone, so there is no data discrepancy in this predictor variable and can be used in the model.

## Converting Quality variables to integer variables

```{r}

convert_to_int_func <- function(data_set_name) {
  
  
  #There are various columns which has below scores, we would convert then to integer values based on the quality.
  
  score <- c('None'=0, 'Po'=1, 'Fa'=2, 'TA'=3, 'Gd'=4, 'Ex'=5)

  data_set_name$PoolQC <- as.integer(revalue(data_set_name$PoolQC, score))
  data_set_name$FireplaceQu <- as.integer(revalue(data_set_name$FireplaceQu, score))
  data_set_name$KitchenQual <- as.integer(revalue(data_set_name$KitchenQual, score))
  data_set_name$ExterQual <- as.integer(revalue(data_set_name$ExterQual, score))
  data_set_name$ExterCond <- as.integer(revalue(data_set_name$ExterCond, score))
  data_set_name$BsmtQual <- as.integer(revalue(data_set_name$BsmtQual, score))
  data_set_name$BsmtCond <- as.integer(revalue(data_set_name$BsmtCond, score))
  data_set_name$GarageQual <- as.integer(revalue(data_set_name$GarageQual, score))
  data_set_name$GarageCond <- as.integer(revalue(data_set_name$GarageCond, score))


  data_set_name$BsmtExposure <- as.integer(revalue(data_set_name$BsmtExposure, 
                                                 c('None'=0, 'No'=1, 'Mn'=2, 'Av'=3, 'Gd'=4)))

  data_set_name$BsmtFinType1 <- as.integer(revalue(data_set_name$BsmtFinType1, 
                                                 c('None'=0, 'Unf'=1, 'LwQ'=2, 'Rec'=3, 'BLQ'=4, 'ALQ'=5, 'GLQ'=6)))
  
  data_set_name$BsmtFinType2 <- as.integer(revalue(data_set_name$BsmtFinType2, 
                                                 c('None'=0, 'Unf'=1, 'LwQ'=2, 'Rec'=3, 'BLQ'=4, 'ALQ'=5, 'GLQ'=6)))

  data_set_name$LotShape <- as.integer(revalue(data_set_name$LotShape, 
                                             c('IR3'=0, 'IR2'=1, 'IR1'=2, 'Reg'=3)))

  data_set_name$GarageFinish<- as.integer(revalue(data_set_name$GarageFinish, 
                                                 c('None'=0, 'Unf'=1, 'RFn'=2, 'Fin'=3)))
  
  data_set_name$MasVnrType <- as.integer(revalue(data_set_name$MasVnrType, 
                                               c('None'=0, 'BrkCmn'=0, 'BrkFace'=1, 'Stone'=2)))
  
  data_set_name$Functional <- as.integer(revalue(data_set_name$Functional, 
                                                 c('Sal'=0, 'Sev'=1, 'Maj2'=2, 'Maj1'=3, 'Mod'=4, 'Min2'=5, 'Min1'=6, 'Typ'=7)))
  

return(data_set_name)
}

complete_data_set <- convert_to_int_func(complete_data_set)

table(sapply(complete_data_set, class))
```

## Feature Engineering 

```{r combining_variables, echo = FALSE}

combining_variables_func <- function(data_set_name) {
  
  #Creating a "total number of bathrooms" variable by adding all bathroom features
  data_set_name$TotalBaths <- data_set_name$BsmtFullBath + 
                                (0.5 * data_set_name$BsmtHalfBath) +
                                data_set_name$FullBath + 
                                (0.5 * data_set_name$HalfBath)
  
  #Creating the age variables 
  data_set_name$RemodelAge <- data_set_name$YrSold - data_set_name$YearRemodAdd
  
  data_set_name$RemodelFlag <- ifelse(data_set_name$RemodelAge == 0, 0,1) #0-No, 1-Yes
  
 data_set_name$TotalAreaSF <- data_set_name$TotalBsmtSF + data_set_name$GrLivArea
 
  #Creating a "total area" variable by adding the basement area and ground living area
 data_set_name$TotalSF <- data_set_name$TotalBsmtSF + 
                            data_set_name$X1stFlrSF + 
                            data_set_name$X2ndFlrSF +
                            data_set_name$LowQualFinSF 


# Creating a "total porch area" variable by adding all porch features
 data_set_name$TotalPorchSF <- data_set_name$OpenPorchSF + 
                                data_set_name$EnclosedPorch + 
                                data_set_name$X3SsnPorch + 
                                data_set_name$ScreenPorch +
                                data_set_name$WoodDeckSF

 # Creating Scores based variables on Quality and Condition features of the dataset

  data_set_name$OverallScore <- data_set_name$OverallQual * data_set_name$OverallCond
  data_set_name$GarageScore <- data_set_name$GarageQual * data_set_name$GarageCond
  data_set_name$ExterScore <- data_set_name$ExterQual * data_set_name$ExterCond
  data_set_name$KitchenScore <- data_set_name$KitchenAbvGr * data_set_name$KitchenQual
  data_set_name$FireplaceScore <- data_set_name$Fireplaces * data_set_name$FireplaceQu
  data_set_name$GarageAreaScore <- data_set_name$GarageArea * data_set_name$GarageQual
  data_set_name$PoolScore <- data_set_name$PoolArea * data_set_name$PoolQC

return(data_set_name)
}

complete_data_set <- combining_variables_func(complete_data_set)

```

## Correlation Plot

```{r plot, fig.height = 12, fig.width = 14}

numeric_train_data <- complete_data_set %>% filter(Set == "Train") %>% 
                                select(which(sapply(.,is.integer)), which(sapply(., is.numeric)))
#, use = "pairwise.complete.obs"
correlation <- cor( numeric_train_data ) %>%
                as.matrix( sort( correlation[,'SalePrice'], decreasing = TRUE )) #sort on decreasing correlations with SalePrice


corrplot.mixed(correlation, tl.col="black", tl.pos = "lt")
```
> From the correlation plot, we found a list of variables that are highly correlated with each other so we can remove one of them.


## Excluding some variables based on correlation plot

```{r}
column_drop_list <- c('Utilities','BsmtFullBath','BsmtHalfBath','FullBath','HalfBath','YearRemodAdd','YrSold',
                      'EnclosedPorch','X3SsnPorch','ScreenPorch','WoodDeckSF','GarageArea','GarageQual','GarageCond',
                      'ExterQual','ExterCond','KitchenQual','FireplaceQu','PoolArea','PoolQC'
                      )

complete_data_set <- complete_data_set[,!(names(complete_data_set) %in% column_drop_list)]
```

##Converting character variables to factors

```{r}
table(sapply(complete_data_set, class))

# List of character variables
variable_classes <- sapply(complete_data_set, class)
names(variable_classes[which(variable_classes=="character")])

```

```{r}

convert_to_factor_func <- function(data_set_name) {
  
#Numeric factors that should be categorical
  data_set_name$MoSold <- factor(data_set_name$MoSold)
  data_set_name$MSSubClass <- factor(data_set_name$MSSubClass)
  
  data_set_name <- data_set_name %>% 
  mutate(Neighborhood = as.factor(Neighborhood),
         Neighborhood2 = fct_collapse(Neighborhood,
                                           "BluesteSawyer" = c("Blueste", "Sawyer"),
                                           "NPkVillMitchel" = c("NPkVill", "Mitchel")))
  
  data_set_name <- data_set_name %>% 
  mutate(LotShape = as.factor(LotShape),
         LotShape2 = fct_collapse(LotShape,
                                       "IR23" = c("IR2", "IR3")))

  data_set_name <- data_set_name %>% 
  mutate(LotConfig = as.factor(LotConfig),
         LotConfig2 = if_else(LotConfig == "CulDSac", 1, 0)) 


#Finding the Variables with character class
  variable_classes <- sapply(data_set_name, class)
  character_variables <- names(variable_classes[which(variable_classes=="character")])

# Converting them to factor
  data_set_name[character_variables] <- lapply(data_set_name[character_variables], factor)

  return(data_set_name)
}

complete_data_set <- convert_to_factor_func(complete_data_set)
```

## Checking for outliers

### Scatterplot to identify the outliers

```{r}
#GrLivArea

ggplot(data=complete_data_set[!is.na(complete_data_set$SalePrice),], aes(x=GrLivArea, y=SalePrice))+
        geom_point() + geom_smooth(method = "lm", se=FALSE, aes(group=1)) +
        scale_y_continuous(breaks= seq(0, 800000, by=100000), labels = comma) +
        geom_text_repel(aes(label = ifelse(complete_data_set$GrLivArea[!is.na(complete_data_set$SalePrice)]>4500, rownames(complete_data_set), '')))

#TotalBsmtSF

ggplot(data=complete_data_set[!is.na(complete_data_set$SalePrice),], aes(x=TotalBsmtSF, y=SalePrice))+
        geom_point() + geom_smooth(method = "lm", se=FALSE, aes(group=1)) +
        scale_y_continuous(breaks= seq(0, 800000, by=100000), labels = comma) +
        geom_text_repel(aes(label = ifelse(complete_data_set$TotalBsmtSF[!is.na(complete_data_set$SalePrice)]>4500, rownames(complete_data_set), '')))

```

### Removing outliers

```{r}

complete_data_set <- complete_data_set[ !(complete_data_set$Id %in% c(524, 1299)), ]

complete_data_set %>% 
  filter(OverallCond == 2 & SalePrice > 12)

complete_data_set <- complete_data_set %>% 
  mutate(OverallCond = if_else(Id == 379, 5L, OverallCond))

```
> Id column is removed from both the datasets, Id is not a predictor.

## Rechecking the Correlation values after cleaning the data

```{r correlation}

train_cleaned <- complete_data_set %>%
                   filter(Set == "Train")

test_cleaned <- complete_data_set %>%
                   filter(Set == "Test") 

#high correlation cor>0.3
for(col in colnames(train_cleaned)){
  if(is.numeric(train_cleaned[,col])){
    if(abs(cor(train_cleaned[,col],train_cleaned$SalePrice)) > 0.5 ){
      print (c(col, abs(cor(train_cleaned[,col],train_cleaned$SalePrice))))
    }
  }
}


# low correaltion cor<0.1
for(col in colnames(train_cleaned)){
  if(is.numeric(train_cleaned[,col])){
    if(abs(cor(train_cleaned[,col],train_cleaned$SalePrice))<0.1){
      print(c(col,cor(train_cleaned[,col],train_cleaned$SalePrice)))
    }
  }
}

```

## Plots to study the important variables to be used in the model

```{r histogram_plots}

#OverallQual

ggplot(data=complete_data_set[!is.na(complete_data_set$SalePrice),], aes(x=factor(OverallQual), y=SalePrice)) +
 geom_boxplot(outlier.colour = "red", outlier.shape = 1)  +
        scale_y_continuous(breaks= seq(0, 800000, by=100000), labels = comma) 

#YearBuilt

ggplot(data=complete_data_set[!is.na(complete_data_set$SalePrice),], aes(x=YearBuilt^15, y=SalePrice))+
        geom_point() + geom_smooth(method = "lm", se=FALSE, aes(group=1)) +
        scale_y_continuous(breaks= seq(0, 800000, by=100000), labels = comma) +
        geom_text_repel(aes(label = ifelse(complete_data_set$YearBuilt[!is.na(complete_data_set$SalePrice)]>4500, rownames(complete_data_set), '')))

ggplot(complete_data_set[!is.na(complete_data_set$SalePrice),], aes(x=Neighborhood, y=SalePrice)) +
        geom_bar(stat='summary', fun.y = "median", fill='blue') +
        theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
        scale_y_continuous(breaks= seq(0, 800000, by=50000), labels = comma) +
        geom_label(stat = "count", aes(label = ..count.., y = ..count..), size=3) +
        geom_hline(yintercept=163000, linetype="dashed", color = "red") #dashed line is median SalePrice
                                      
                                          
hist.data.frame(complete_data_set %>%
  select("GrLivArea","RemodelAge"))

hist.data.frame(complete_data_set %>%
  select("GarageFinish","GarageCars","GarageAreaScore"))

hist.data.frame(complete_data_set %>%
  select("OverallScore","ExterScore","FireplaceScore","KitchenScore"))

hist.data.frame(complete_data_set %>%
  select("HeatingQC","CentralAir","FireplaceScore"))

hist.data.frame(complete_data_set %>%
  select("BsmtQual","TotalBsmtSF","BsmtFinSF1","TotalPorchSF","KitchenAbvGr"))

hist.data.frame(complete_data_set %>%
  select("X1stFlrSF","TotRmsAbvGrd","TotalBaths","TotalSF","OpenPorchSF"))



```

## Linear Regression Model

```{r}
tc <- trainControl(method = "cv", number = 5)

set.seed(123)

(lm_model <- train(log(SalePrice+1) ~ (factor(OverallQual)+factor(BsmtQual)*TotalBsmtSF+YearBuilt^20+log(X1stFlrSF+1)*TotRmsAbvGrd
                                        +factor(Neighborhood)+GarageFinish^3+GarageCars*TotalBaths+RemodelAge+
                                          TotalSF^2+OverallScore+ExterScore*MSZoning+
                                          factor(FireplaceScore)+Fireplaces+GarageAreaScore*log(GrLivArea+1)^14+
                                          HeatingQC^4+CentralAir^2+TotalPorchSF+BsmtFinSF1*MasVnrArea^2
                                        +KitchenScore*KitchenAbvGr^2+OverallQual+OpenPorchSF+TotalAreaSF),
                     data = select(train_cleaned, -Id),
                     preProcess = c("center","scale"),
                     trControl = tc,
                     method = "lm"))
summary(lm_model)



```

## Predictions

```{r}

predictions <- data.frame(Id = test_cleaned$Id,
                          lm = exp(predict(lm_model, newdata= test_cleaned))
                          )

head(predictions)
```


## In-sample RMSE value

```{r}
rmse <- function(actual, predicted) sqrt(mean((actual - predicted)^2))

rmse(train_cleaned$SalePrice, exp(predict(lm_model, newdata= train_cleaned)) )
```

## in-sample R-squared value

```{r}
r2 <- function(actual, predicted){
  TSS <- sum((actual - mean(actual))^2)
  RSS <- sum((actual - predicted)^2)
  1 - RSS/TSS
}

r2(train_cleaned$SalePrice, exp(predict(lm_model, newdata= train_cleaned)))
```

## Predicted vs actual SalePrice plot

```{r}

plot(exp(predict(lm_model, newdata= train_cleaned)),                                # Draw plot using Base R
     train_cleaned$SalePrice,
     xlab = "Predicted Values",
     ylab = "Observed Values")
abline(a = 0,                                        # Add straight line
       b = 1,
       col = "red",
       lwd = 2)

```

## Submission

```{r}
#kaggle score of 0.13                         
submit <- test_cleaned %>% 
  select(Id) %>% 
  mutate(SalePrice = exp(predict(object = lm_model, newdata = test_cleaned)))

write.csv(submit, "kaggle_submit.csv", row.names = F )
head(submit)
```


